\begin{intro}
  In the previous chapter, we studied discretizations with
  $\div V_h = Q_h$ with two advantages. First, due to
  \slideref{Corollary}{galerkin-mixed-u-kerb} the velocity error is
  independent of the pressure. Second, the divergence converges faster
  than the gradient. A natural question arising is whether we can do
  something similar for the Stokes problem. There, the equation
  \begin{gather*}
    \form(\div v_h, q_h) = 0 \qquad\forall q_h\in Q_h,
  \end{gather*}
  would immediately imply $\div v_h=0$, that is, the discrete solution
  is exactly divergence free.
  
  The answer to this question is a current research topic. So far,
  beginning with the element by Scott and Vogelius, several methods
  have been proposed for special mesh geometry or macro meshes. The
  difficulty is balancing the condition $\div V_h = Q_h$ with the
  $H^1$-conformity of the velocity space. All the spaces in the
  previous chapter were only $\Hdiv$-conforming with discontinuous
  tangential components.
  
  A fairly simple solution to this question though can be obtained by
  using discontinuous Galerkin methods. These were introduced to
  obtain formulations \emph{consistent} with $H^1$ while not
  \emph{conforming}. Thus, we can apply them directly to
  Raviart-Thomas and Brezzi-Douglas-Marini elements to obtain a
  consistent method with divergence free solutions.

  We begin this chapter by a quick review of the interior penalty
  method before diving into divergence conforming methods.
\end{intro}

\section{The interior penalty method}
\input{../fem/ip}
\subsection{Bounded formulation in $H^1$}
\input{../fem/ip-lifting}

\section{Divergence conforming IP}
\begin{remark}
  The extension of the interior penalty method to vector-valued
  problems is obvious. Furthermore, since the method generates an
  elliptic bilinear form on the discontinuous space $V_h$, this
  ellipticity is inherited by any subspace of
  $V_h\cap\Hdiv(\domain)$. Thus, we can write down the weak
  formulation of a divergence conforming DG method for the Stokes
  equations. In the following definition, we assume slip or no-slip
  boundary conditions, that is, $v\cdot\n=0$ on the whole boundary.
\end{remark}

\begin{Definition}{hdiv-ip}
  A divergence conforming DG method for the Stokes equations consists
  of a discrete velocity space $V_h\subset \Hdiv_0(\domain)$ and a
  pressure space $Q_h\subset L^2_0(\domain)$ such that
  \begin{gather}
    \label{eq:hdivdg:1}
    \div V_h = Q_h.
  \end{gather}
  Using the interior penalty bilinear form $a_h(.,.)$, we search for
  solutions $(u_h,p_h)\in V_h\times Q_h$ such that for all $(v,q)\in
  V_h\times Q_h$ there holds
  \begin{gather}
    \label{eq:hdivdg:2}
    a_h(u_h,v) +\form(\div v,p_h)+\form(\div u_h,q) = f(v).
  \end{gather}
\end{Definition}

\begin{remark}
  Due to the fact that $V_h\not\subset V$, we have introduced the  norm
  $\norm{.}_{1,h}$ on $V_h$. In particular, the norm $\norm{.}_1$ is
  not defined for all elements of $V_h$. Therefore, we need a
  modification of Fortin's lemma (\slideref{Lemma}{fortin}), where the
  norm on the left hand side of the stability
  estimate~\eqref{eq:galerkin:16} uses the discrete norm, namely,
  \begin{gather*}
    \norm{\Pi_{V_h}v}_{V_h} \le c \norm{v}_V,
  \end{gather*}
\end{remark}

\begin{Lemma}{dg-fortin}
  Let $\{\mesh_h\}$ be a shape-regular sequence of meshes. Then, the
  \putindex{canonical interpolation} operators of the
  Brezzi-Douglas-Marini and Raviart-Thomas elements admit the bound
  \begin{gather}
    \label{eq:hdivdg:4}
    \norm{I_h v}_{1,h} \le c \snorm{v}_1
  \end{gather}
\end{Lemma}

\begin{proof}
  First, we note that all degrees of freedom are defined as cell or
  face integrals with smooth weight functions. Thus, they are bounded
  on $H^1$. Thus, since the local polynomial spaces are finite
  dimensional, there holds on the reference cell $\refcell$ and its
  faces $\refface$:
  \begin{align*}
    \norm{I_{\refcell} v}_{1;\refcell} &\le c \snorm{v}_{1;\refcell},
    \\
    \norm{I_{\refcell} v}_{0;\refface} &\le c \snorm{v}_{1;\refcell}.   
  \end{align*}
  On shape regular meshes, we have the scaling property
  \begin{align*}
    \snorm{f}_{m;\cell} &\simeq h_\cell^{\frac{d}{2}-m},
    \\
    \snorm{f}_{m;\face} &\simeq h_\face^{\frac{d-1}{2}-m},
  \end{align*}
  such that for a face $\face$ of cell $\cell$
  \begin{align*}
    \norm{I_{\cell} v}_{1;\cell} & \le c \snorm{v}_{1;\cell},
    \\
    \norm{I_{\cell} v}_{0;\refface} &\le c h^{\frac12} \snorm{v}_{1;\cell}.
  \end{align*}
  We conclude
  \begin{gather*}
    \norm{I_h v}_{1,h}^2 \le \sum_{\cell\in\mesh_h}
    \biggl[\norm{I_{\cell} v}_{1;\cell}^2
    + 4 \sum_{\face\subset\d\cell}
    \norm*{\tfrac{\ipp_0}{h_F} I_{\cell} v}_{0;\face}^2
    \biggr] \le c \snorm{v}_{1}^2.
  \end{gather*}
\end{proof}

\begin{Corollary}{hdivdg-infsup}
  Assume that the inf-sup condition~\eqref{eq:stokes:1} in
  \slideref{Theorem}{stokes-infsup} holds.
  Then, the method in \slideref{Definition}{hdiv-ip}
  admits the inf-sup condition
  \begin{gather}
    \label{eq:hdivdg:3}
    \inf_{q_h\in Q_h} \sup_{v_h\in V_h}
    \frac{\form(\div v_h,q_h)}{\norm{v_h}_{1,h}\norm{q_h}_0} \ge \beta,
  \end{gather}
  with a constant $\beta >0$ independent of $h$.
\end{Corollary}

\begin{proof}
  First, we make use of the fact that $q_h\in Q_h \subset Q$ to deduce
  from \slideref{Theorem}{stokes-infsup} the there is a function $w\in
  V$ with $\div v=q_h$ and $\norm{v}_1 \le \norm{q_h}_0$. To this
  function, we apply the \putindex{Fortin operator} to define $v_h =
  I_h v$. By the preceding lemma, we have
  \begin{gather*}
    \norm{v_h}_{1,h} \le c \norm{v}_1 \le \norm{q_h}_0,
  \end{gather*}
  which proves the inf-sup condition.
\end{proof}

\begin{Theorem}{hdivdg-convergence}
  Assume that $(u_h,p_h)\in V_h\times Q_h$ is the solution to the
  divergence conforming DG method in \slideref{Definition}{hdiv-ip} and
  that the continuous Stokes problem is well-posed as in
  \slideref{Theorem}{stokes-infsup}. Then, for the Raviart-Thomas
  pairs $RT_k/\P_k$ and $RT_{[k]}/\Q_k$ with $k\ge 1$ and
  $u$ sufficiently smooth there holds
  \begin{align}
    \label{eq:hdivdg:5}
    \norm{u-u_h}_{1,h} &\le h^k \snorm{u}_{k+1}, \\
    \label{eq:hdivdg:7}
    \norm{p-p_h}_{0} &\le h^k \bigl(\snorm{u}_{k+1} + \snorm{p}_{k}\bigr).
  \end{align}
  Furthermore,
  \begin{gather}
    \label{eq:hdivdg:6}
    \div u_h = 0.
  \end{gather}
\end{Theorem}

\begin{proof}
  The proof follows the lines of the abstract theory of
  \slideref{Theorem}{galerkin-mixed-u-kerbh} and
  \slideref{Theorem}{galerkin-mixed-p}. But since the setting with
  $V_h\not\subset V$ exceeds the assumptions of the abstract theory,
  we adapt the proofs instead of using the results.

  Due to consistency of the method, we have
  \begin{gather*}
    a_h(u-u_h, v_h) + \form(\div v_h, p-p_h)
    + \form(\div u-\div u_h, q_h) = 0.
  \end{gather*}
  Testing with $v_h=0$ and using $\div V_h = Q_h$ immediately yields
  $\div u_h = \div u = 0$, or
  \begin{gather*}
    \ker{B_h} \subset \ker B.
  \end{gather*}
  In order to use the ellipticity of $a_h(.,.)$, we insert
  arbitrary functions $w_h \in \ker{B_h}$ and $r_h\in Q_h$. Choosing
  $q_h = 0$ yields the error equation
  \begin{gather}
    \label{eq:hdivdg:8}
    a_h(u_h-w_h, v_h) + \form(\div v_h, p_h-r_h)
    = a_h(u-w_h, v_h) + \form(\div v_h, p-r_h).
  \end{gather}
  Testing with $v_h=u_h-w_h$ and employing $\div v_h=0$, we obtain
  \begin{gather*}
    \ellipa \norm{u_h-w_h}_{1,h}^2
    \le a_h(u_h-w_h,u_h-w_h)
    = a_h(u-w_h,u_h-w_h).
  \end{gather*}
  Now, we use the canonical interpolation $w_h = I_h u$ to obtain
  \begin{gather*}
    \ellipa \norm{u_h-w_h}_{1,h}^2
    \le \frac\ellipa2 \norm{u_h-w_h}_{1,h}^2
    + \frac{c}{2\ellipa} h^{2k} \snorm{u}_{k+1}^2.
  \end{gather*}

  Finally, we use the inf-sup condition to find a test function
  $v_h\in V_h$ such that $\div v_h = p_h-r_h$ and $\beta \norm{v_h}_{1,h}
  \le \norm{p_h-r_h}$. Then, the error equation~\eqref{eq:hdivdg:8}
  yields
  \begin{multline*}
    \norm{p_h-r_h}
    = \frac{\form(\div v_h,p_h-r_h)}{\norm{p_h-r_h}}
    \\
    = \frac{a_h(u-u_h, v_h) + \form(\div v_h, p-r_h)}
    {\norm{p_h-r_h}}
    \le \tfrac{\norm{a_h}}\beta \norm{u-u_h}_{1,h}
    + \norm{p-r_h}_0
    .
  \end{multline*}
  Using the previously proven error estimate for $u_h$ and the
  $L^2$-projection $r_h = \Pi_h p$ yields the result.
\end{proof}

\section{Error estimates by duality}

\begin{intro}
  So far, we have only considered estimates in the so called
  \putindex{energy norm}, that is, a norm such that $a_h(.,.)$ is
  bounded and elliptic\footnote{We use the term energy norm loosely
    here. Strictly speaking, the energy norm would be
    $\norm{v}_A = \sqrt{a_h(v,v)}$.}.

  In the context of elliptic equations, we have seen the duality
  argument of Aubin and Nitsche, which allows us to obtain optimal
  estimates in weaker norms, for instance in $L^2$.

  A particular difficulty here is the fact, that we have to test the
  dual solution with the error \emph{and} exploit some kind of
  Galerkin orthogonality. Thus, we cannot use consistency as before
  and will introduce residual operators later. The analysis here is a
  simplified version of the corresponding results
  in~\cite{GiraultKanschatRiviere14}.
\end{intro}

\begin{Definition}{dual-stokes}
  The \putindex{dual problem} to the Stokes problem in weak for
  consists of finding $(u^*,p^*)\in V_h\times Q_h$ such that for all
  $v\in V$ and $q\in Q$ there holds
  \begin{gather}
    \label{eq:hdivdg:9}
    \form(\nabla v,\nabla u^*) + \form(\div u^*,q) + \form(\div v,p^*)
    = \form(\psi,v).
  \end{gather}
\end{Definition}

\begin{Assumption}{stokes-regularity}
  The dual Stokes problem admits the elliptic regularity estimate
  \begin{gather}
    \label{eq:hdivdg:10}
    \norm{u^*}_{2} \le c \norm{f}_0.
  \end{gather}
\end{Assumption}

\begin{remark}
  Like for scalar elliptic equations, the elliptic regularity
  assumption holds for domains with smooth boundary or with piecewise
  smooth boundary where every corner is convex.
\end{remark}

\begin{Definition}{hdivdg-residual-operators}
  For the solutions $(u,p)\in V\times Q$ and $(u^*,p^*)\in V\times Q$
  of the primal and dual Stokes problem, respectively, we define the
  residual operators
  \begin{align}
    \operatorname{Res}(u,p;v) &= a_h(u,v)+\form(\div v,p) - \form(f,v),
    \\
    \operatorname{Res}^*(v;u^*,p^*)
                              &= a_h(v,u^*)+\form(\div v,p^*) -
                                \form(\psi,v),
  \end{align}
  for $v\in V+V_h$.
\end{Definition}

% From Girault/Kanschat/Riviere

\begin{Lemma}{hdivdg-residual-1}
  Let $(u,p)\in V\times Q$ be the solution to the Stokes problem with
  right hand side $f\in L^2(\domain;\R^d)$. Assume $u\in
  H^s(\domain;\R^d)$ and $p\in H^{s-1}(\domain)$ with $s>3/2$. Then,
  we have for $v\in V+V_h$:
  \begin{gather}
    \label{eq:hdivdg:11}
    \form(f,v) = \form(\nabla u,\nabla v)_{\mesh_h}
    -\forme(\nabla u,\mvl{v\otimes n})_{\faces_h^i}
    -\forme(\d_n u,v)_{\faces_h^\d}
    + \form(\div v,p).
  \end{gather}
\end{Lemma}

\begin{proof}
  We set out from the strong form of the Stokes equations and
  integrate by parts.
  \begin{align*}
    \form(f,v) &= \form(-\Delta u + \nabla p, v)
    \\
    &= \form(\nabla u,\nabla v)_{\mesh_h}
      - \sum_{\cell\in\mesh_h} \forme(\d_n u,v)_{\d\cell}
      - \form(\div v,p)
      .
  \end{align*}
  Under the regularity assumptions of the lemma, all of these
  integrals make sense at least as duality pairings. In particular,
  $\d_n u\in L^2(\d\cell)$, and thus we can split $\d\cell$ into
  individual faces. Therefore,
  \begin{gather*}
    \sum_{\cell\in\mesh_h} \forme(\d_n u,v)_{\d\cell}
    = \forme(\nabla u,\mvl{v\otimes n})_{\faces_h^i}
    +\forme(\d_n u,v)_{\faces_h^\d}.
  \end{gather*}
  The proof concludes by collecting the results.
\end{proof}

\begin{Corollary}{hdivdg-residual-2}
  The residual operators can be expressed as
  \begin{gather}
    \label{eq:hdivdg:12}
    \begin{split}
    \operatorname{Res}(u,p;v)
    &= a_h(u,v)
      - \form(\nabla u,\nabla v)_{\mesh_h}
      \\
      &\qquad
      + \forme(\nabla u,\mvl{v\otimes n})_{\faces_h^i}
      +\forme(\d_n u,v)_{\faces_h^\d}.
      \\
    \operatorname{Res}^*(u^*,p^*;v)
    &= a_h(v,u^*)
      - \form(\nabla u,\nabla v)_{\mesh_h}
      \\
      &\qquad
      + \forme(\nabla u,\mvl{v\otimes n})_{\faces_h^i}
      +\forme(\d_n u,v)_{\faces_h^\d}.      
    \end{split}
  \end{gather}
  In particular, the residual operators do not depend on the pressure
  solutions.
\end{Corollary}

\begin{Theorem}{hdivdg-l2}
  Let the assumptions of \slideref{Theorem}{hdivdg-convergence}
  and \slideref{Assumption}{stokes-regularity} hold. Then,
  \begin{gather}
    \norm{u-u_h}_0 \le c h^{k+1} \snorm{u}_{k+1}.
  \end{gather}
\end{Theorem}

\begin{Problem}{hdivdg-l2}
  Adapt the proof of \slideref{Theorem}{ip-lifting-l2} to prove
  \slideref{Theorem}{hdivdg-l2}.
\begin{solution}
   We again consider the auxiliary problem
  \begin{gather*}
    a(v,u^*) +(\nabla\cdot v, p^*)= \form(u-u_h,v),\qquad\forall v\in V.
  \end{gather*}
  Using the definition of the dual residual, we obtain the equation
  \begin{gather*}
    \form(u-u_h, v) = a_h(v,u^*) +(\nabla\cdot v, p^*)- \Res^*(u^*,p^*;v),\qquad\forall v\in V+V_h.
  \end{gather*}
  Testing with $v=u-u_h$ yields
  \begin{gather*}
    \norm{u-u_h}^2 = a_h(u-u_h,u^*) +(\nabla\cdot (u-u_h), p^*)- \Res^*(u^*, p^*;u-u^*).
  \end{gather*}
  Additionally, we us the error equation
  \begin{gather*}
    a_h(u-u_h, v_h)+(\nabla\cdot v_h, p-p_h) = \Res(u,p;v_h),
  \end{gather*}
  tested with $v_h = I_h u^*$, to obtain
  \begin{align*}
    \norm{u-u_h}^2 &= a_h(u-u_h,u^*-I_h u^*) +(\nabla\cdot (u-u_h), p^*-q_h)\\
    &\quad- \Res^*(u^*, p^*;u-u^*)
    -\underbrace{(\nabla\cdot I_h u^*, p-p_h)}_{0} + \Res(u,p;I_h u^*).    
  \end{align*}
  Using the regularity of $u^*$, the first term on the right
  admits the estimate
  \begin{gather*}
    \abs{a_h(u-u_h, u^*-I_h u^*)}
    \le \norm{u-u_h}_{1,h}\norm{u^*-I_hu^*}_{1,h}
    \le c h \norm{u-u_h}_{1,h}.
  \end{gather*}
  as before.
  
  For the second term consider $q_h=\Pi_{Q_h} p^*$
  \begin{align*}
   (\nabla\cdot (u-u_h), p^*-\Pi_{Q_h} p^*)\leq \norm{\nabla \cdot  (u-u_h)} \norm{p^*-\Pi_{Q_h} p^*}=0
  \end{align*}
  
  For the third term we use \slideref{Lemma}{ip-lifting-residual-2}
  to obtain
  \begin{gather*}
    \abs{\Res^*(u^*,u-u_h)} \le c h \snorm{u^*}_2 \norm{u-u_h}_{1,h}.
  \end{gather*}
  Finally, using $\jmp{u^*} = 0$, the same lemma yields
  \begin{align*}
    \abs{\Res(u, I_h u^*)}
    &\le c h \snorm{u}_2
      \bigl(\norm{\sqrt{\ipp_h}\jmp{I_h u^*}}_{\faces_h^i}
      + \norm{\sqrt{\ipp_h}I_h u^*}_{\faces_h^\d}\bigr)
    \\
    & = c h \snorm{u}_2
      \bigl(\norm{\sqrt{\ipp_h}\jmp{u^*-I_h u^*}}_{\faces_h^i}
      + \norm{\sqrt{\ipp_h}(u^*-I_h u^*)}_{\faces_h^\d}\bigr)
    \\
    & \le c h \snorm{u}_2 h^k \snorm{u^*}_{k+1}
  \end{align*}
  This is exactly the same proof we used before.
  Using the energy estimate in \slideref{Theorem}{ip-lifting-h1} we
  can conclude the prove.
\end{solution}

\end{Problem}

%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End: 
