\subsection{The macroelement technique}

\begin{intro}
  The arguments of the previous section can be extended to the case
  where instead of individual mesh cell $\cell$ we condider small
  groups of such cells, called
  \putindex{macroelement}s~\cite{Stenberg84}. This will give us some
  more freedom for testing the inf-sup condition locally, but possibly
  not on each cell separately. In order to get uniform estimates, we
  introduce a generalization of the concept of shape regularity.
\end{intro}

\begin{Definition}{macro-equivalence}
  A \define{macroelement} $M\subset \T_h$ is a union of cell
  $\cell_i\in\T_h$. Given the mappings $\Phi_\cell\colon \widehat\cell \to
  \cell$, there is a reference macroelement $\widehat M$ and a mapping
  $\Phi_M\colon \widehat M\to M$ such that $\Phi_M(\widehat M) = M$. We say
  that $M$ is equivalent to $\widehat M$.
\end{Definition}

\begin{Definition}{macro-spaces}
  For a \putindex{macroelement} $M$, we introduce the spaces
  \begin{align}
    \label{eq:macroelement:27}
    V_M &= \bigl\{ u\in H^1_0(M;\R^d) \big\vert
            \;\exists v_h\in V_h\colon u=v_{h|M}\bigr\},
    \\
    \label{eq:macroelement:28}
    Q_M &= \bigl\{ p\in L^2_0(M) \big\vert
            \;\exists q_h\in Q_h\colon p=q_{h|M}\bigr\},
  \end{align}
  the kernel of the discrete, local gradient operator
  \begin{gather}
    \label{eq:macroelement:29}
    \ker{B^T_M} = \bigl\{q\in Q_M \big\vert 
    \;\forall v\in V_M\colon \form(\div v,q) = 0\bigr\}.
  \end{gather}
  For a mesh $\mesh_h$ which is subdivided into a nonoverlapping set
  of macros $M_j$, we define the space of pressure functions
  $Q_M^0\subset Q$ which are constant on each macro with the
  $L^2$-projection $\Pi_M\colon Q_h\to Q_M^0$.
\end{Definition}

\begin{Lemma}{macro-local}
  Let $\{M\}$ with $M\subset \mesh_h$ be a set of macroelements
  equivalent to the same reference macroelement $\widehat M$. Let the
  family $\{\mesh_h\}$ be shape regular and assume
  $\ker{B^T_{M}} = \{0\}$. Then, there is a constant
  $\beta_M>0$ independent of $h$ such that for all $M$ there holds
  \begin{gather}
    \label{eq:macroelement:31}
    \inf_{p\in Q_M} \sup_{v\in V_M}
    \frac{\form(\div v,q)_M}{\norm{v}_{1,M}\norm{q}_{0,M}}
    \ge \beta_{\widehat M}.
  \end{gather}
\end{Lemma}

\begin{proof}
  For each $M$, the existence of such a constant $\beta_{M}$ is
  equivalent to the fact that the smallest nonzero singular value is
  strictly greater than zero in finite dimensional spaces. Next, we
  observe that $\nabla \Phi_{M}$ is bounded by the
  $\nabla\Phi_{\cell_i}$, which in turn is uniformly bounded by the
  shape regularity. Therefore, by a cell-wise scaling argument and the
  fact that $\widehat M$ has only finitely many cells we have
  \begin{gather}
    \beta_{\widehat M} = \min_M \beta_M > 0.
  \end{gather}
\end{proof}

\begin{remark}
  Depending on the technique of proof being used, we also may decide
  to impose~\eqref{eq:macroelement:31} directly for each macroelement.
\end{remark}

\begin{Lemma}{macro-projection}
  Let $\mesh_h$ be subdivided into a nonoverlapping set of
  macroelements $M$ such that there is a finite set of reference macro
  elements $\widehat M$, and let $\ker{B^T_M} = \{0\}$. Then, there is
  a constant $c_1>0$ such that for all $q_h\in Q_h$ exists $v_h\in V_h$
  satisfying
  \begin{gather}
    \label{eq:macroelement:32}
    \begin{split}
      \form(\div v_h, q_h) = \form(\div v_h,{(\identity-\Pi_M)} q_h)
      &\ge c_1 \norm{(\identity-\Pi_M) q_h}_0^2\\
      \norm{v_h}_2 & \le \norm{(\identity-\Pi_M) q_h}_0.
    \end{split}
  \end{gather}
\end{Lemma}

\begin{proof}
  For any $q_h\in Q_h$, we have $(\identity-\Pi_M) q_h\in Q_M$ for each
  macroelement $M$. Applying \slideref{Lemma}{macro-local}, for each
  $M$ there exists $v_M\in V_M$ such that
  \begin{gather}
    \form(\div v_M, {(\identity-\Pi_M)} q_h)_M
    \ge c_1 \norm{v_M}_{1,M}\norm{(\identity-\Pi_M) q_h}_{0,M},
  \end{gather}
  with $c=\min \beta_{\widehat M}$. Let now $v_M$ such that
  $\norm{v_M}_{1,M} = \norm{(\identity-\Pi_M) q_h}_{0,M}$ and $v$ such that
  $v_{|M} = v_M$. Since $v_M$ is zero on $\d M$, we have $v\in
  V_h$. By construction,~\eqref{eq:macroelement:32} holds.
\end{proof}

\begin{Lemma}{macro-constant}
  Let $\{M_j\}$ be a covering of $\mesh_h$ with macroelements. For any
  two macros $M_i$ and $M_j$ let $T_{ij}$ be their common
  boundary. Then, if every $T_{ij}$ of codimension one contains at
  least one degree of freedom controlling $u\cdot\n$, there is a
  constant $c_2>0$ such that for every $q_h\in Q_h$ exists $w_h\in
  V_h$ such that
  \begin{gather}
    \label{eq:macroelement:33}
    \begin{split}
      \form(\div w_h, \Pi_M q_h) &\ge c_2 \norm{\Pi_M q_h}^2_0\\
      \norm{w_h}_1 &\le \norm{\Pi_M q_h}_0.
    \end{split}
  \end{gather}
\end{Lemma}

\begin{Problem}
  Prove \slideref{Lemma}{macro-constant}
\end{Problem}

\begin{Theorem}{macroelement}
  Let the assumptions of \slideref{Lemma}{macro-projection}
  hold. Then, the pair $V_h\times Q_h$ is inf-sup stable.
\end{Theorem}

\begin{proof}
  Let $q_h\in Q_h$. Then, applying \slideref{Lemma}{macro-projection}
  and \slideref{Lemma}{macro-constant}, we obtain functions $v_h$ and
  $w_h$ such that for $u_h = v_h+ \delta w_h$ for any $\delta>0$ there
  holds
  \begin{align}
    \form(\div u_h, q_h)
    &= \form(\div v_h, q_h) + \form(\div w_h, q_h) \\
    &\ge c_1 \norm{(\identity-\Pi_M) q_h}_0^2
      + \delta c_2 \norm{\Pi_M q_h}_0^2
      - \delta \norm{w_h}_1 \norm{(\identity-\Pi_M) q_h}_0 \\
    & = (c_1 - \delta) \norm{(\identity-\Pi_M) q_h}_0^2
      + \delta c_2 \norm{\Pi_M q_h}_0^2.
  \end{align}
  Choosing $\delta = c_1/(1+c_2)$, we obtain
  \begin{gather}
    \label{eq:macroelement:34}
    \begin{split}
      \form(\div u_h, q_h) &\ge \frac{c_1c_2}{1+c_2} \norm{q}_0^2 \\
      \norm{u}_1 &\le \norm{q}_0.
    \end{split}
  \end{gather}
\end{proof}


%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:
