\input{./fig/tikzsettings}
\maketitle

\section*{Preface}
%\input{../preface}

These notes compile information from various sources, which should be
consulted for more details on particular subjects. These are in
particular the books by Daniele Boffi, Franco Brezzi, and Michel
Fortin~\cite{BoffiBrezziFortin13} and by Alexandre Ern and Jean-Luc
Guermond~\cite{ErnGuermond04} on the general theory of mixed finite
elements, the books by Dietrich Braess~\cite{Braess97,Braess13} and
Philippe Ciarlet~\cite{Ciarlet88} on elasticity, the book by Vivette
Girault and Pièrre-Arnaud Raviart~\cite{GiraultRaviart86} on
incompressible fluids, the book by Peter Monk~\cite{Monk03}, and the
works by Douglas Arnold, Richard Falk, and Ragnar
Winther~\cite{ArnoldFalkWinther06acta,ArnoldFalkWinther10} on the
finite element exterior calculus.

Special thanks go to Dr.~Daniel Arndt for contributing problems,
solutions, and improving the notes at many points. Furthermore, I
thank Vladislav Olkhovskiy and Saurabh Mehta for valuable corrections
and additions.
\thispagestyle{empty}
\setcounter{page}{0}


\tableofcontents

\chapter{Introduction: from elliptic to mixed problems}
We begin our course of mixed finite element methods by studying a
vector-valued elliptic problem. Then, we study its dependence on its
parameters and naturally arrive at a mixed formulation. We derive a
few properties of mixed systems and then turn our attention to the
first example: the Stokes equations of incompressible flow. We close
the chapter by considering mixed systems as first order conditions
for constrained minimization problems.

\input{mixed/mixedintro}

\section{Abstract saddle-point systems}
\input{mixed/abstract-saddle}

\section{Relation to constrained minimization}
\input{mixed/lagrange}

\section{Additional problems and experiments}

\begin{Problem}{elasticity-locking-convergence}
  The graph for $\lambda=100$ in \cref{fig:elasticity-locking}
  suggests the following behavior: on very coarse meshes, the solution
  is badly approximated. Eventually, the approximation cathes up and
  is even faster than orer $h^{-1}$ until it approximates
  asymptotically the straight line for $\lambda=1$. The curve for
  $\lambda=1000$ seems to show the beginning of this behavior.

  \begin{enumerate}
  \item Compute on finer meshes to see if this is actually true. If
    such a behavior is confirmed, this implies that the part of the
    bilinear form with $\lambda$ is less important on finer meshes,
    something which cannot be covered by Céa's lemma from the elliptic
    toolbox.
  \item Compute the energy norm of the error as in the symmetric
    finite element toolbox. What do those curves look like? There, the
    constant in Céa's lemma is 1.
  \item Can we draw any conclusions on the best approximation in the
    energy norm?
  \end{enumerate}
\end{Problem}

\chapter{Conditions for well-posedness}
\label{sec:mixed-wellposedness}
In this chapter, we will first modify conditions for well-posedness in
finite dimensions from positive definiteness to the general case. In
particular, we will derive a quantitative formulation, which we will
study for infinite dimensional problems in the second section. In the
third section, we derive the inf-sup condition for mixed problems as a
special case.
\input{mixed/infsup}

\chapter{The Stokes problem}
\label{cha:stokes}
\input{mixed/stokes}

\chapter{Mixed formulation of elliptic problems}
\label{cha:darcy}
\input{mixed/darcy}

\chapter{Divergence conforming discontinuous Galerkin methods}
\label{cha:hdivdg}
\input{mixed/hdivdg}

\chapter{Maxwell's equations and the de Rham complex}
\section{Maxwell's equations}
\label{cha:maxwell}
\input{mixed/maxwell}

\section{The de Rham complex}
\label{cha:derham}
\input{mixed/derham}

\appendix
\chapter{Appendix}
\section{Notation}

\input{mixed/notation}

\section{Useful tools}
\input{mixed/tools}

\section{Derivation of the Lamé-Navier equations}
\label{sec:lame-navier}
\input{lame-navier}

\bibliographystyle{apalike}
\bibliography{all}
\printindex


%%% Local Variables: 
%%% mode: latex
%%% TeX-master: "main"
%%% End: 
