\documentclass[a4paper,10pt]{article}
\usepackage[utf8]{inputenc}
\usepackage[margin=1.4in]{geometry}

\setlength{\parindent}{0pt}
\begin{document}

\section*{\centering List of relevant topics for the oral examination on ``Finite Element Methods for Flow Simulations''}

\textit{Part A} will be mandatory as starting point for the
examination. For the main part, the candidate can select \textit{Part B} or \textit{Part C}. Here is an overview:
\begin{itemize}
\item The short \textit{Part A} of the examination will focus on the modeling of flows, covers Section 1-2, and serves
as starting point.
\item \textit{Part B} will be concerned with the analysis and numerics of linear flow problems. It covers
Section 3-9 of the script.
\item The remaining \textit{Part C} will be devoted to the analysis and numerics of nonlinear flow and covers
Section 10-13 of the script.
\end{itemize}

\textbf{Part A:} \textit{Modeling of flows and basis equations of fluid dynamics}
\begin{itemize}
\item \textbf{Section 1:} \textit{Modeling of flows}\\
Lagrangian and Eulerian representation. Velocity. Transport theorem (no proof). Conservation of
mass and of momentum (no proofs). Role of constitutive laws.
\item \textbf{Section 2:} \textit{Basic equations of fluid dynamics}\\
Incompressible Navier-Stokes, Euler and Stokes models. Typical boundary conditions for Navier-Stokes model. 
Role of non-dimensionalization.
\end{itemize}

\textbf{Part B:} Analysis and numerics of linear(ized) flow problems
\begin{itemize}
\item \textbf{Section 3:} \textit{Theory and FEM for scalar problems}\\
Rotation-free incompressible flows. Pressure Poisson model. Well-posedness of diffusion model
with Dirichlet resp. Neumann boundary conditions (result, sketch of proof). Well-posedness of
Ritz-Galerkin approximation (result, sketch of proof). Error estimates for FEM (results).
\item \textbf{Section 4:} \textit{Theory of stationary Stokes problem}\\
Variational form of mixed and constrained problems. Well-posedness of Stokes problem (result
and sketch of proof, but without derivation of inf-sup condition). Generalization to saddle point
problems (result).
\item \textbf{Section 5:} \textit{Conforming FEM for stationary Stokes problem}\\
Galerkin problem. Necessary solvability condition. Well-posedness and error estimates (results,
sketch of proof for velocity error). FE pairs with discrete inf-sup condition.
\item \textbf{Section 6:} \textit{Time-dependent Stokes problem}\\
Basics of vector-valued functions. Main theorem of linear evolution problems (a-priori estimates,
well-posedness result, rough sketch of proof). Application to linear transient mixed problems, e.g.
to Stokes problem.
\item \textbf{Section 7:} \textit{Discretization of time-dependent Stokes problem}\\
Spatial semidiscretization of linear transient mixed problems. (Generalized) Stokes projector. 
Velocity error estimates (result, sketch of proof). Fully-coupled temporal discretization (backward
Euler). Problem of Uzawa approach. Helmholtz decomposition. Idea of velocity-pressure splitting.
\item \textbf{Section 8:} \textit{Generalized Oseen problem}\\
Motivation of Oseen problem. Variational formulation and role of convective term. Well-posedness
of constrained problem. FE-discretization with inf-sup stable FE-pairs. Velocity error estimates
(sketch of proof). Div-div stabilization.
\item \textbf{Section 9:} \textit{Time-dependent Oseen problem}\\
Variational formulation of continuous problem as mixed and constrained problems. Well-posedness
(result and sketch of proof). Semidiscretization in space. Error estimates for velocity (result, sketch
of proof with modified Stokes projector).
\end{itemize}

\textbf{Part C:} Analysis and numerics of nonlinear flow problems
\begin{itemize}
\item \textbf{Section 10:} \textit{Stationary incompressible Navier-Stokes problem}\\
Variational formulation of continuous problem as mixed and constrained nonlinear problems.
Solvability of continuous problem (result and sketch of proof via sequence of Galerkin problems
and Brouwer’s fixed point theorem). Uniqueness in small data case (result and sketch of proof).
\item \textbf{Section 11:} \textit{Numerical solution of stationary incompressible Navier-Stokes problem}\\
Galerkin approximation. A-priori estimate (result and proof). Idea of Petrov-Galerkin methods.
Solution techniques for nonlinear algebraic problems.
\item \textbf{Section 12:} \textit{Theory of time-dependent incompressible Navier-Stokes problem}\\
Evolution formulation of continuous problem. Solvability of constraint evolution problem (result
and sketch of proof via Galerkin approximation). Uniqueness (results and open problems).
\item \textbf{Section 13:} \textit{Numerical solution of transient incompressible Navier-Stokes problem}\\
Spatial semidiscretization. Existence and a-priori estimates of semidiscrete solutions (result and
sketch of proof). Convergence of semidiscrete solutions: Stokes projection. Idea of estimate of
discretization error. Temporal discretization with $\Theta$-scheme.
\end{itemize}
\end{document}
