
% \begin{intro}
%   In this section we extend the weakening of continuity, which we
%   explored for boundary values in Section~\ref{sec:nitsches-method}
%   using Nitsche's method to interior interfaces between mesh
%   cells. While the methods obtained may look much more complicated,
%   the mathematical analysis is completely analogue to that
%   section. Thus, we can be fairly brief.
% \end{intro}

\begin{intro}
  We review the basic definitions necessary to describe discontinuous
  Galerkin (DG) methods. In particular, we need the sets of faces
  $\F_h$ of a mesh, discontinuous piecewise polynomial spaces and
  broken integrals.
\end{intro}

\begin{Definition}{dg-faces}
  Let $\T_h$ be a mesh of $\Omega \subset \R^d$ consisting of mesh
  cells $T_i$. For every boundary facet $F\subset \partial T_i$, we
  assume\footnote{This assumption can indeed be relaxed} that either
  $F \subset \partial \Omega$ or $F$ is a boundary facet of another
  cell $T_j$. In the second case, we indicate this relation by
  labeling this facet $F_{ij}$. The set of all facets $F_{ij}$ is the
  set of interior faces $\F_h^i$. The set of facets on the boundary is
  $\F_h^\partial$.
\end{Definition}

\begin{Definition}{dg-spaces}
  The discontinuous finite element space on $\T_h$ is constructed by
  concatenation of all shape function spaces $P_T$ for $T\in \T_h$
  without additional continuity requirements:
  \begin{gather}
    V_h = \bigl\{v\in L^2(\Omega) \big|
    v_{|T} \in P_T \;\forall T\in \T_h\bigr\}.
  \end{gather}
\end{Definition}

\begin{Definition}{broken-integrals}
  For any set of cells $\mesh_h$ or faces $\faces_h$, we define the bilinear
  forms
  \begin{align}
    \form(u,v)_{\mesh_h} &= \sum_{\cell\in\mesh_h} \form(u,v)_\cell, \\
    \forme(u,v)_{\faces_h} &= \sum_{\face\in\faces_h} \forme(u,v)_\face. \\
  \end{align}
\end{Definition}

\begin{intro}
  We start out with the equation
  \begin{gather*}
    -\Delta u = f.
  \end{gather*}
  Integrating by parts on each mesh cell yields
  \begin{gather*}
    \form(-\Delta u,v )_\cell
    = \form(\nabla u, \nabla v)_\cell - \forme(\d_n u, v)_{\d\cell} = \form(f,v)_T.
  \end{gather*}
  We realize that the choice of discontinuous finite element spaces
  introduces a consistency term on the interfaces between cells and on
  the boundary.

  On interior faces, there is the issue that $u$ and
  $\d_n u$ actually have two values on the interface, one from the
  left cell and one from the right. Therefore, we have to consolidate
  these two values into one. To this end, we introduce the concept of
  a numerical flux, which constructs a single value out of these
  two. Thus, we introduce on the interface $\face$ between two cells
  $\cell^+$ and $\cell^-$
  \begin{gather*}
    \mathcal F(\nabla u) = \frac{\nabla u^+ + \nabla u^-}{2} = :
    \mvl{\nabla u}.
  \end{gather*}

  Using $\forme(\d_n u,v) = \forme(\nabla u,v\n)$ we change our point
  of view and instead of integrating over the boundary $\d\cell$, we
  integrate over a face $\face$ between two cells $\cell^+$ and
  $\cell^-$. Adding up integrals from both sides, we obtain the term
  \begin{gather*}
    -\forme(\mvl{\nabla u},v^+\n^+ +v^-\n^-)_{\face}
    = -2\forme(\mvl{\nabla u},\mvl{v\n})_{\face}.
  \end{gather*}
  On boundary faces, we simply get
  \begin{gather*}
    \forme(\d_\n u,v)_{\face}.
  \end{gather*}

  Adding over all cells and faces, we obtain the equation
  \begin{gather*}
    \form(\nabla u,\nabla v)_{\T_h}
    -2\forme(\mvl{\nabla u},\mvl{v\n})_{\F_h^i}
    -\forme(\d_\n u,v)_{\F_h^\d} = \form(f,v)_{\domain}.
  \end{gather*}

  Following the idea of Nitsche, we symmetrize this term
  to obtain
  \begin{multline*}
    \form(\nabla u,\nabla v)_{\T_h}
    -2\forme(\mvl{\nabla u},\mvl{v\n})_{\F_h^i}
    -2\forme(\mvl{u\n},\mvl{\nabla v})_{\F_h^i}
    \\
    -\forme(\d_\n u,v)_{\F_h^\d}
    -\forme(u,\d_\n v)_{\F_h^\d}
    = \form(f,v)_{\domain}
    - \forme(u^o,\d_n v)_{\F_h^\d}.
  \end{multline*}
  Here the second term on the right was introduced for consistency.
  Finally, it turns out that this method is not stable and needs
  stabilization by a jump term. This will be done in
  \blockref{Definition}{ip}. Before, we introduce the notation for
  averaging and jump operators.
\end{intro}

\begin{Notation}{dg-operators}
  Let $\face$ be a face between the cells $\cell^+$ and $\cell^-$. Let
  $\n^+$ and $\n^-=-\n^+$ be the outer normal vectors of the cells at a
  point $x\in \face$. For a function $u\in V_h$, the traces $u^+$ and
  $u_-$ of $u$ on $\face$ taken from the cell $\cell^+$
  and $\cell^-$ are defined as:
  \begin{align*}
    u^+(x) &= \lim_{\epsilon\searrow 0} u(x-\epsilon\n^+), \\
    u^-(x) &= \lim_{\epsilon\searrow 0} u(x-\epsilon\n^-).
  \end{align*}
  We define the \define{averaging operator} $\mvl{.}$ and the
  \define{jump operator} $\jmp{.}$ as
  \begin{gather}
    \label{eq:ip:1}
    \mvl{u} = \frac{u^++u^-}{2},
    \qquad
    \jmp{u} = u^+-u^-.
  \end{gather}
  Not that the sign of the jump of $u$ depends on the choice of the
  cells $\cell^+$ and $\cell^-$. It will only be used in quadratic
  terms.
\end{Notation}

\begin{remark}
  The jump can be denoted as the mean value of the product of a
  function and the normal vector,
  \begin{gather}
    \jmp{u} = 2\mvl{u\n}\cdot\n^+ = -2\mvl{u\n}\cdot\n^-.
  \end{gather}
\end{remark}

\begin{Definition}{ip}
  The \define{interior penalty method}\footnote{Also known as
    symmetric interior penalty (SIPG) or IP-DG.} uses the bilinear
  form
  \begin{multline}
    \label{eq:ip:2}
    a_h(u,v) = \form(\nabla u,\nabla v)_{\mesh_h}
    + \forme(\ipp_h\jmp{u},\jmp{v})_{\faces_h^i}
    + \forme(\ipp_h u,v)_{\faces_h^\d}
    \\
    -2\forme(\mvl{\nabla u},\mvl{v\n})_{\faces_h^i}
    -2\forme(\mvl{u\n},\mvl{\nabla v})_{\faces_h^i}
    \\
    - \forme(\d_n u,v)_{\faces_h^\d}
    - \forme(u,\d_n v)_{\faces_h^\d},
  \end{multline}
  and the linear form
  \begin{gather}
    \label{eq:ip:3}
    f_h(v) = \form(f,v)_{\domain} - \forme(u^D,\d_n v)_{\faces_h^\d}
    + \forme(\ipp_h u,v)_{\faces_h^\d},
  \end{gather}
  where $f$ is the right hand side of the equation and $u^D$ the
  Dirichlet boundary value.
\end{Definition}

\begin{Definition}{ip-norm}
  On the space $V_h$ we define the norm $\norm{.}_{1,h}$ by
  \begin{gather}
    \label{eq:ip:4}
    \norm{v}_{1,h}^2 = \sum_{\cell\in\mesh_h} \norm{\nabla v}_\cell^2
    + \sum_{\face\in\faces_h^i} \norm{\sqrt{\ipp_h}\jmp{v}}_\face^2
    + \sum_{\face\in\faces_h^\d} \norm{\sqrt{\ipp_h}v}_\face^2.
  \end{gather}
\end{Definition}

\begin{Problem}{ip-norm}
  Prove that the norm defined in (\ref{eq:ip:4}) is indeed a norm on $V_h$.
\begin{solution}
\begin{align*}
  0=\norm{v}_{1,h}^2 &= \sum_{\cell\in\mesh_h} \norm{\nabla v}_\cell^2
    + \sum_{\face\in\faces_h^i} \norm{\sigma_h\jmp{v}}_\face^2
    + \sum_{\face\in\faces_h^\d} \norm{\sigma_hv}_\face^2
\end{align*}
implies first of all $v|_T \equiv const.$ for all $T \in \mathbb{T}_h$.
Furthermore, $\norm{\sigma_h\jmp{v}}_\face^2=0$ implies $v \equiv const.$
and the last condition gives $v \equiv 0$.

 We use the trace inequality
 \begin{align*}
  \norm{v}_{0,\partial T}^2
  \lesssim \left( h_T^{-1} \norm{v}_{0,T}^2+\norm{v}_{0,T}\norm{\nabla v}_{0,T}\right)
  \quad \forall v \in H^1(T).
 \end{align*}
 Then, we can estimate
 \begin{align*}
  &\sum_{\face\in\faces_h^i} \norm{\sigma_h\jmp{v}}_\face^2 + \sum_{\face\in\faces_h^\d} \norm{\sigma_hv}_\face^2\\
  &\lesssim \min_{F\in F_h^i\cup F_h^\partial} \sigma_{h,F} \sum_{T \in \mathbb{T}_h}\norm{v}_{0,\partial T}^2 \\
  &\lesssim \min_{F\in F_h^i\cup F_h^\partial} \sigma_{h,F} \sum_{T \in \mathbb{T}_h}\left( h_T^{-1}
  \norm{v}_{0,T}^2+\norm{v}_{0,T}\norm{\nabla v}_{0,T}\right)\\
  &\lesssim \min_{F\in F_h^i\cup F_h^\partial} \sigma_{h,F} \sum_{T \in \mathbb{T}_h}\left( h_T^{-1}
  \norm{v}_{0,T}^2+h_T\norm{\nabla v}_{0,T}\right)
 \end{align*}
 and thus
 \begin{align*}
    \norm{v}_{1,h}^2 &= \sum_{\cell\in\mesh_h} \norm{\nabla v}_\cell^2
    + \sum_{\face\in\faces_h^i} \norm{\sigma_h\jmp{v}}_\face^2
    + \sum_{\face\in\faces_h^\d} \norm{\sigma_hv}_\face^2\\
    &\lesssim \min_{F\in F_h^i\cup F_h^\partial} \sigma_{h,F}
      \sum_{T \in \mathbb{T}_h}\left( h_T^{-1} \norm{v}_{0,T}^2+\norm{\nabla v}_{0,T}\right)\\
  &\lesssim \norm{v}_1^2
  \end{align*}
\end{solution}
\end{Problem}

\begin{Lemma}{ip-stability}
  Let $\T_h$ be shape-regular and chosen on each face $\face$ as
  $\sigma_h = \sigma_0/h_\face$, where $h_T$ is the minimal diameter
  of a cell adjacent to $\face$. Then, there is a $\sigma_0>0$ such
  that there exists a constant $\ellipa>0$, such that independent of
  $h$ there holds
  \begin{gather}
    \label{eq:ip:5}
    a_h(u_h,u_h) \ge \ellipa \norm{u_h}_{1,h}^2 \quad \forall u_h\in V_h.
  \end{gather}
\end{Lemma}

\begin{Problem}{ip-stability}
  Prove \blockref{Lemma}{ip-stability}.
\begin{solution}
We first note the estimate
\begin{align*}
 (\boldsymbol{n}\cdot\nabla v_h)_e^2
 &\leq C (h_K^{-1}\norm{\nabla v_h}_{0,K}^2+\norm{\nabla v_h}_{0,K}\norm{\nabla^2 v_h}_{0,K}) \\
 &\leq C \left(\frac{1}{h_K}+\frac{p_K^2}{h_K}\right) \norm{\nabla v_h}_{0,K}^2\\
 &\leq C \frac{p_K^2}{h_K} \norm{\nabla v_h}_{0,K}^2
 = C\frac{\sigma_h}{\delta} \norm{\nabla v_h}_{0,K}^2
\end{align*}

Testing the bilinear form symmetrically, we obtain
\begin{align*}
 a_h(u_h,u_h) &= \norm{\nabla u_h}_0^2
    + \ipp_h\norm{\jmp{u_h}}_{\faces_h^i}
    + \ipp_h\norm{u_h}_{\faces_h^\d}
    \\
    &-4\forme(\mvl{\nabla u_h},\mvl{u_h\n})_{\faces_h^i}
    -2 \forme(\d_n u_h,u_h)_{\faces_h^\d}.
\end{align*}
and the last two terms can be estimated by
\begin{align*}
 \forme(\mvl{\nabla u_h},\mvl{u_h\n})_{\faces_h^i} &=\forme(n^+\cdot\nabla u_h^+-n^-\cdot\nabla u_h^-, u_h^+-u_h^-)\\
 &\leq \frac{\epsilon}{2}\norm{n^+\{\{\nabla u_h\}\}}_0^2+\frac{1}{2\epsilon}\norm{[[u_h]]}_0^2\\
 &\leq C\frac{\sigma_h\epsilon}{2\delta}\norm{\nabla u_h}_{0,K}^2+\frac{1}{2\epsilon}\norm{[[u_h]]}_0^2
\end{align*}
and therefore
\begin{align*}
 a_h(u_h,u_h) -\gamma \norm{u_h}_{1,h}^2&\geq \norm{\nabla u_h}_0^2 \left(1-\gamma-C\frac{\sigma_h\epsilon}{2\delta}\right)\\&\quad
    + \ipp_h\norm{\jmp{u_h}}_{\faces_h^i}\left(1-\gamma-\frac{1}{2\epsilon\sigma_h}\right).
\end{align*}

Hence, we have to choose $\epsilon, \delta>0$ such that
\begin{align*}
 1-\gamma-C\frac{\sigma_h\epsilon}{2\delta}>0 \\
 1-\gamma-\frac{1}{2\epsilon\sigma_h}>0 .
\end{align*}
This is possible for all $\gamma\in(0,1)$ and in fact we get the lower limits
\begin{align*}
\epsilon&>\frac{1}{2(1-\gamma)\sigma_h}\\
 \delta&>C \frac{\sigma_h \epsilon}{2(1-\gamma)}.
\end{align*}


\end{solution}

\end{Problem}

\begin{Lemma}{ip-consistence}
  Let $f\in L^2(\domain)$ and let the boundary conditions admit that
  for the solution to
  \begin{xalignat*}2
    -\Delta u &= f &\text{in }&\domain, \\
    u &= u^D &\text{on }&\d\domain,
  \end{xalignat*}
  there holds $u\in H^{1+\epsilon}(\domain)$ for a positive
  $\epsilon$. Then, the interior penalty method is consistent, that
  is,
  \begin{gather}
    a_h(u,v_h) = f_h(v_h)\quad\forall v_h\in V_h.
  \end{gather}
\end{Lemma}

\begin{proof}
  From $f\in L^2(\domain)$ we deduce that
  $\nabla u\in \Hdiv(\domain)$. Thus, with the extra regularity, the
  traces of $\d_n u$ on faces are well-defined and coincide from both
  sides. The remainder is integration by parts.
\end{proof}

\begin{Theorem}{ip-convergence}
  For $k\ge 1$ let $\P_k\subset P_\cell$ and $u\in H^{s+1}(\domain)$ with
  $1/2 \le s \le k$. Then, the interior penalty method admits the
  error estimate
  \begin{gather}
    \norm{u-u_h}_{1,h} \le c h^s \snorm{u}_{s+1}.
  \end{gather}
  If furthermore the boundary condition admits \putindex{elliptic
    regularity},
there holds
  \begin{gather}
    \norm{u-u_h}_{0} \le c h^{s+1} \snorm{u}_{s+1}.
  \end{gather}
\end{Theorem}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "main"
%%% End:




